---
title: "MAT 5101 - Problem Set 1"
subtitle: "Group 2"
author: "Erik Celdran, Jade Rosales, Marc Valeros"
date: "2024-02-10"
output: pdf_document
header-includes: |
  \usepackage{graphicx}
  \graphicspath{ {figures/} }
  \usepackage{array}
  \usepackage{amsmath}
  \usepackage{xcolor}
  \usepackage{bigints}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Problem Set 1

1. Prove that $lim_{x\rightarrow -1}{2x}+1=-1.$

Note: For any given $\epsilon > 0$, there exists a $\delta > 0$ such that $|f (x) - f (x_0)| < \epsilon$ whenever $|x - x_0| < \delta$.

Draft:

$$|2x+1 - (-1)| < \epsilon$$
$$|2x+2| < \epsilon$$
$$|2||x-(-1)|<\epsilon$$
$$ |x-(-1)| < \frac{\epsilon}{|2|} $$
$$ \delta \leq \frac{\epsilon}{2} $$ 

Solution:
Given $\epsilon>0$, let $\delta \leq \frac{\epsilon}{2}$, we show that when $|x-(-1)|<\delta$, then $|2 x+1-(-1)|<\epsilon$.
$$
\begin{gathered}
|x-(-1)|<\delta \\
|x-(-1)|<\frac{\epsilon}{2} \\
|2||x-(-1)|<\epsilon \\
|2 x+2|<\epsilon \\
|2 x+1-(-1)|<\epsilon
\end{gathered}
$$

Which is what we wanted to show. Thus, $\lim _{x \rightarrow 1} 2 x+1=-1$.

2. Determine all the numbers $c$ which satisfy the conclusions of the Mean Value Theorem for the following function and graph using $\mathrm{R}$ with the point/s identified. $f(x)=x^3-4 x^2-2 x-5$ on $[-10,10]$.

Note: The Mean Value Theorem states that for a continuous and differentiable function $f(x)$ on the interval $[a, b]$ there exists such number $c$ from the interval $(a, b)$, that $f^{\prime}(c)=\frac{f(b)-f(a)}{b-a}$.

First, evaluate the function at the endpoints of the interval:
$$
\begin{aligned}
& f(10)=575 \\
& f(-10)=-1385
\end{aligned}
$$

Next, find the derivative: $f^{\prime}(c)=3 c^2-8 c-2$ 

Calculate $\mathrm{f}(\mathrm{a})$ and $\mathrm{f}(\mathrm{b})$ when $\mathrm{a}=-10, \mathrm{~b}=10$.
$$
\mathrm{a}=-10
$$
$$
\begin{gathered}
f(a)=a^3-4 a^2-2 a-5 \\
f(-10)=(-10)^3-4(-10)^2-2(-10)-5 \\
=-1000-400+20-5 \\
=f(a)=-1385
\end{gathered}
$$
$$
\mathrm{b}=10
$$
$$
\begin{gathered}
f(b)=b^3-4 b^2-2 b-5 \\
f(10)=10^3-4(10)^2-2(10)-5 \\
=1000-400-20-5 \\
=f(b)=575
\end{gathered}
$$

Calculate:
$$
\begin{gathered}
f^{\prime}(\xi)=\frac{f(b)-f(a)}{b-a} \\
3 \xi^2-8 \xi-2=\frac{575-(-1385)}{10-(-10)} \\
3 \xi^2-8 \xi-2=98 \\
3 \xi^2-8 \xi-100=0
\end{gathered}
$$

Form the equation: $3 c^2-8 c-2=\frac{(575)-(-1385)}{(10)-(-10)}$
Simplify: $3 c^2-8 c-2=98$

Solve the equation on the given interval: $c=\frac{4}{3}-\frac{2 \sqrt{79}}{3}, c=\frac{2(2+\sqrt{79})}{3}$

Answer: $\frac{4}{3}-\frac{2 \sqrt{79}}{3} \approx-4.59212961154373, \frac{2(2+\sqrt{79})}{3} \approx 7.25879627821039$

3. Find the point $c$ that satisfies the mean value theorem for integrals on the interval $[-1,1]$. The function is $f(x)=2 e^x$.

Solution: The function is a polynomial and is therefore continuous.
From the integral mean value theorem, we have:
$$
\int_b^a f(x) d x=f(\epsilon)(b-a)
$$

Substituting the values in the formula, we get:
$$
\begin{gathered}
\int_{-1}^1 2 e^x d x=2 e^\epsilon(1-(-1)) \\
\left.2 e^x\right]_{-1}^1=2\left(2 e^\epsilon\right) \\
2 e-\frac{2}{e}=4 e^\epsilon \\
\ln (\frac{2 e-\frac{2}{e}}{4})=\ln (e^\epsilon) \\
\epsilon=\ln \frac{2 e-\frac{2}{e}}{4} \\
\epsilon \approx 0.1614394
\end{gathered}
$$
```{r}
# exp(1) is equivalent to e
epsilon = ln((2*exp(1)-2*exp(-1))/4)
epsilon
```R

4. Consider the function $f(x)=\cos (x / 2)$.
a Find the fourth Taylor polynomial for $f$ at $x=\pi$.
b Use the fourth Taylor polynomial to approximate $\cos (\pi / 2)$.
c Use the fourth Taylor polynomial to bound the error.

a. For $f(x)=\cos (x / 2)$, the values of the function and its first three derivatives at $x=\pi$ are as follows:
$$
\begin{aligned}
& f(x)=\cos (x / 2) \rightarrow f(\pi)= 0\\
& f^{\prime}(x)= -\sin \left(\frac{x}{2}\right) \frac{1}{2} \rightarrow f^{\prime}(\pi)= -\frac{1}{2}\\
& f^{\prime \prime}(x)=-\frac{1}{4} \cos \left(\frac{x}{2}\right) \rightarrow f^{\prime \prime}(\pi)= 0\\
& f^{\prime \prime \prime}(x)=\frac{1}{8} \sin \left(\frac{x}{2}\right) \rightarrow f^{\prime \prime \prime}(\pi)= \frac{1}{8}\\
& f^{\prime \prime \prime \prime}(x)=\frac{1}{16} \cos \left(\frac{x}{2}\right) \rightarrow f^{\prime \prime \prime \prime}(\pi)= 0\\
\end{aligned}
$$

Thus, the fourth Taylor polynomials at $x=\pi$ are given by
$$
\begin{aligned}
& p_4=f(\pi)+f^{\prime}(\pi)(x-\pi)+\frac{f^{\prime \prime}(\pi)}{2 !}(x-\pi)^2+\frac{f^{\prime \prime \prime}(\pi)}{3 !}(x-\pi)^3+\frac{f^{\prime \prime \prime \prime}(\pi)}{4 !}(x-\pi)^4=-\frac{1}{2}(x-\pi)+\frac{1}{48}(x-\pi)^3
\end{aligned}
$$

b. Solution. Using the fourth Taylor polynomial at $x= \pi$, we can estimate
$$
\cos (\pi / 2) \approx p_4(\pi)=0-\frac{1}{2}(\pi - \pi)+0+\frac{1}{48}(\pi-\pi)^3= 0
$$

c. For $R_4(\pi)$ :
$$
p_4=0-\frac{1}{2}(\pi - \pi)+0+\frac{1}{48}(\pi-\pi)^3= 0
$$
f4d=function(x){(1/16)*(cos(x/2))}

f4d(c(pi,pi))

[1] 3.826895e^-18 3.826895e^-18

For $R_4(\pi)$ :
$$
(3.826895e^{-18}) / 24 *(\pi-\pi)^ 4
$$
[1] 0
$$
0-\frac{1}{2}(\pi - \pi)+0+\frac{1}{48}(\pi-\pi)^3= 0
$$
[1] 0


Since $f^{\prime \prime \prime\prime}(x)=\frac{1}{16} \cos \left(\frac{x}{2}\right)$, the maximum value of $f^{\prime \prime \prime \prime}$ on the interval $(\pi,\pi)$ is $f^{\prime \prime \prime \prime}(\pi) \approx 0$. Therefore, we have
$$
R_4(\pi) \leq \frac{0}{4 !}(\pi-\pi)^34 \approx 0
$$

5. If $f l(x)$ is the machine approximated number of a real number $x$ and $\epsilon$ is the corresponding relative error, then show that $f l(x)=(1-\epsilon) x$.

if $\epsilon$ is the corresponding relative error then, $E(x_A) = x - x_A$, where $x_A$ is $fl(x)$

$$E_r(x_A) = \frac{E(x_A)}{x}$$

$$E_r(x_A) = \frac{x - fl(x)}{x}$$

$$\epsilon = E_r(x_A)$$

Which shows that:

$$\epsilon = \frac{x - fl(x)}{x}$$

$$\epsilon x = x - fl(x)$$

$$f l(x) = x-\epsilon x$$

$$f l(x) = (1 - \epsilon)x$$


6. For the following numbers $x$ and their corresponding approximations $x_A$, find the number of significant digits in $x_A$ with respect to $x$ and find the relative error.
a. $x=451.01, x_A=451.023$
b. $x=-0.04518, x_A=-0.045113$
c. $x=23.4604, x_A=23.4213$

Note:

Relative error

$$E(x_A) = x - x_A$$
$$E_r(x_A) = \frac{E(x_A)}{x}$$

Significant Digits

$$\left|x-x_A\right| \leq \frac{1}{2} \beta^{s-r+1}$$

Draft:

a. $x=451.01, x_A=451.023$

Relative Error:
\begin{center}
$E(x_a) = 451.01 - 451.023 = -0.013$

$E_r(x_a) = \frac{-0.013}{451.01} = -0.00002882419$
\end{center}
Significant Digits:
\begin{center}
$|x-x_A| = 0.013 < 0.05 = \frac{1}{2}\times10^-1$. But $10^2 < 451.01$. Therefore $s = 2$ and $r = 4$

$0.013 \leq \frac{1}{2} \beta^{2-4+1} = \frac{1}{2} \times 10^{-1} = 0.0005$
\end{center}
b. $x=-0.04518, x_A=-0.045113$

Relative Error:
\begin{center}
$E(x_a) = -0.04518 - (-0.045113) = -0.000067$

$E_r(x_a) = \frac{-0.000067}{-0.04518} = 0.00148295706$
\end{center}
Significant Digits:
\begin{center}
$|x-x_A| = 0.000067 < 0.0005 = \frac{1}{2}\times10^{-3}$. But $10^{-2} < 0.04518$. Therefore $s = -2$ and $r = 2$

$0.000067 \leq \frac{1}{2} \beta^{-2-2+1} = \frac{1}{2} \times 10^{-3} = 0.0005$
\end{center}
c. $x=23.4604, x_A=23.4213$

Relative Error:
\begin{center}
$E(x_a) = 23.4604 - 23.4213 = 0.0391$

$E_r(x_a) = \frac{0.0391}{23.4604} = 0.00166663824$
\end{center}
Significant Digits:
\begin{center}
$|x-x_A| = 0.0391 < 0.05 = \frac{1}{2}\times10^{-1}$. But $10^{1} < 23.4213$. Therefore $s = 1$ and $r = 3$

$0.0391 \leq \frac{1}{2} \beta^{1-3+1} = \frac{1}{2} \times 10^{-1} = 0.05$
\end{center}
7. Find the condition number for the following functions
a. $f(x)=2 x^2$
b. $f(x)=2 \pi^x$
c. $f(x)=2 b^x$

Note: The condition number of a function $f$ at a point $x = c$ is given by
$$\left|\frac{f^{\prime}(c)}{f(c)} c\right|$$

Draft:

a.

$f(x)=2 x^2$, for all $x \epsilon R$

Then $f'(x)=4 x$, so that

$$\left|\frac{f^{\prime}(x)}{f(x)} x\right| = \left|\frac{4 x}{2 x^2} x\right| = \left|\frac{4}{2}\right| = 2$$

b.

$f(x)=2 \pi^x$, for all $x \epsilon R$

Then $f'(x)=2 \pi^{x}\ln(\pi)$, so that

$$\left|\frac{f^{\prime}(x)}{f(x)} x\right| = \left|\frac{2 \pi^{x}\ln(\pi)}{2 \pi^x} x\right| = \left|x\right|{\ln(\pi)}$$

c.

$f(x)=2 b^x$, for all $x \epsilon R$

Then $f'(x)=2 b^x\ln(b)$, so that

$$\left|\frac{f^{\prime}(x)}{f(x)} x\right| = \left|\frac{2 b^{x}\ln(b)}{2 b^x} x\right| = \left|x\right|{\ln(b)}$$

8. Determine if the following series converges or diverges. If it converges determine its sum.
$$ \sum_{n=1}^{\infty} \frac{1}{2^n} $$
Divergence test:
$$
\lim _{n \rightarrow \infty} \frac{1}{2^n}=\frac{1}{\infty}=0
$$

The series is either convergent or divergent since the limit is 0 as $\mathrm{n}$ approaches $\infty$, it means.
Listing the sequence gives us $\frac{1}{2}, \frac{1}{4}, \frac{1}{8}, \frac{1}{16},\frac{1}{32},\frac{1}{64}, \ldots, \frac{1}{2^n}$
Given $a_1=\frac{1}{2}$ and $r=\frac{1}{2}$, solve $S_n$
$$
\begin{gathered}
S_n=\frac{a_1\left(1-r^n\right)}{1-r} \\
=\frac{\frac{1}{2}\left(1-\frac{1}{2}^n\right)}{1-\frac{1}{2}} \\
=\frac{\frac{1}{2}\left(1-\frac{1}{2}^n\right)}{\frac{1}{2}} \\
=1-\frac{1}{2}^n
\end{gathered}
$$

Now, the summation of the series $\frac{1}{2^n}$ starting at 1 as $\mathrm{n}$ approaches $\infty$ is equal to the limit of $S_n$ as $\mathrm{n}$ approaches $\infty$.
Solve for $\lim _{n \rightarrow \infty} S_n=1-\frac{1}{2}^n$
$$
\begin{gathered}
\lim _{n \rightarrow \infty} S_n=1-\frac{1}{2}^n \\
=1-\frac{1}{2}^{\infty} \\
=1-0 \\
=1
\end{gathered}
$$

So, the series is convergent and its sum is 1 .
