---
title: "MAT 5101 - Problem Set 2"
subtitle: "Group 2"
author: "Erik Celdran, Jade Rosales, Marc Valeros"
date: "2024-02-10"
output: pdf_document
header-includes: |
  \usepackage{graphicx}
  \graphicspath{ {figures/} }
  \usepackage{array}
  \usepackage{amsmath}
  \usepackage{xcolor}
  \usepackage{bigints}
  \graphicspath{ {./} }
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(PolynomF)
library(matrixcalc)
library(pracma)
```
## Problem Set 2

1. Use the bisection method with a hand calculator or computer to find the real root of $x^3-x-3=0$. Use an error tolerance of $\epsilon=0.0001$. Graph the function $f(x)=x^3-x-3$ and label the root.

```{r}
# Define the function
f <- function(x) { x^3 - x - 3 }
```

```{r}
# Set the initial interval and error tolerance, since f(a) < 0 , f(b) > 0
a <- 1
b <- 2
eps <- 0.0001

#initialize data frame 

df = data.frame(matrix(ncol=7))
colnames(df) = c("#","a","b","c","f(b)","f(c)","error (b-a)"); 
ctr = 0

bisect = function(f,a,b,eps){
  while(TRUE){
    ctr <<- ctr + 1
    c = (a+b)/2
    df[nrow(df)+1,] <<- list(ctr,a,b,c,f(b),f(c),b-a)
    if (abs(b-c)<=eps) {df <<- df[-1,]; return(c)}
    if (f(b)*f(c)<=0) {a=c} else {b=c}
  }
}

root = bisect(f,a,b,eps)
print(df, row.names = FALSE)
root # bisection result

```

```{r echo=TRUE}
curve(f, xlim=c(-2,2), ylim=c(-5,5), col="blue", lwd=3)
abline(h = 0, lty = 2)
abline(v = 0, lty = 2)
points(root, 0, pch=19, col="black")
text(root, 0, "x = 1.671692", pos = 3)
```


2. The function $f(x)=-3 x^3+2 e^\frac{x^2}{2}-1$ has values of zero near $x=-0.5$ and $x=0.5$.
a. What is the derivative of $f$ ?
$$f'(x)=-9x^2+2 xe^\frac{x^2}{2}$$
b. If you begin Newton's method at $x=0$, which root is reached? How many iterations to achieve an error less than $10^{-5}$ ?

Newton's Iterative Formula:
$x_{n+1}=x_n-\frac{f\left(x_n\right)}{f^{\prime}\left(x_n\right)}, \quad n \geq 0$

```{r}
#define function
f=function(x){-3*x^3+2*exp(x^2/2)-1}
df=function(x){-9*x^2+2*x*exp(x^2/2)}

x=0

x1=x-(f(x)/df(x))
x1
```

We encounter an issue because the results returns undefined.

c. Begin Newton's method at another starting point to get the other zero.

We begin Newton's method at $x=0.5$

```{r}
x=0.5

x1=x-(f(x)/df(x))
x1

x2=x1-(f(x1)/df(x1))
x2

x3=x2-(f(x2)/df(x2))
x3

x4=x3-(f(x3)/df(x3))
x4

x5=x4-(f(x4)/df(x4))
x5

x6=x5-(f(x5)/df(x5))
x6

error=abs(x6-x5)
error
```

5 iterations are needed at $x=0.5$ to achieve an error less than $10^{-5}$. 

3. Use the function from no. 2 and find the root using the secant method where $x_0=0$ and $x_1=1$. Use an error tolerance of $\epsilon=0.001$.

Secant Method:
$$x_{n+1}=x_n-f\left(x_n\right) \cdot \frac{x_n-x_{n-1}}{f\left(x_n\right)-f\left(x_{n-1}\right)}, \quad n \geq 1$$

```{r}
f=function(x){-3*x^3+2*exp(x^2/2)-1}
sm=function(a,b){a-f(a)*(a-b)/(f(a)-f(b))}

x0=0
x1=1

x2=sm(x1,x0)
x2

x3=sm(x2,x1)
x3

x4=sm(x3,x2)
x4

x5=sm(x4,x3)
x5

x6=sm(x5,x4)
x6

x7=sm(x6,x5)
x7

error=abs(x7-x6)
error
```

4. Consider the system
$$
\begin{gathered}
10.2 x+2.4 y-4.5 z=14.067 \\
-2.3 x-7.7 y+11.1 z=-0.996 \\
-5.5 x-3.2 x+0.9 z=-12.645
\end{gathered}
$$
a. Present the augmented matrix of the system.
```{r}
# Define the coefficients matrix (A)
A <- matrix(c(10.2, 2.4, -4.5, -2.3, -7.7, 11.1, -5.5, -3.2, 0.9), nrow = 3, byrow = TRUE)

# Define the constants vector (b)
b <- c(14.067, -0.996, -12.645)

# Create the augmented matrix
Ab <- cbind(A, b)

# Print the augmented matrix
cat("Augmented Matrix:\n")
print(Ab)
```

b. Solve the system using $A x=L U_X=L y=b$ and round the final answer to 4 decimal digits.

```{r}
# Perform LU decomposition
LU <- lu(A)

# Solve the system using forward substitution (L y = b)
y <- solve(LU$L, b)

# Solve the system using backward substitution (U x = y)
x <- solve(LU$U, y)

# Round the solution to 4 decimal places
x <- round(x, digits = 4)

# Print the solution
cat("\nSolution:\n")
print(x)
```


c. Find the residual vector if the correct solution is $x=1.4531001, y=-1.5891949, z=-0.2748947$.
```{r}
# Define the solution vector
sol <- c(1.4531001, -1.5891949, -0.2748947)

# Calculate the residual vector (b - Ax)
residual <- b - A %*% sol

# Print the residual vector
cat("\nResidual Vector:\n")
print(residual)
```

Therefore, the solution to the system is:

x = 1.4531 (rounded to 4 decimal places)
y = -1.5892 (rounded to 4 decimal places)
z = -0.2749 (rounded to 4 decimal places)

5. Compute the Frobenius norm, maximum column sum, and maximum row sum of the matrix:
$$
\left(\begin{array}{ccc}
10.2 & 2.4 & 4.5 \\
-2.3 & 7.7 & 11.1 \\
-5.5 & -3.2 & 0.9
\end{array}\right)
$$
```{r}
# Define the matrix
mat <- matrix(c(10.2, 2.4, 4.5,
               -2.3, 7.7, 11.1,
               -5.5, -3.2, 0.9), nrow = 3, byrow = TRUE)

# Compute Frobenius norm
frobenius_norm <- sqrt(sum(mat^2))

# Compute maximum column sum
max_col_sum <- max(colSums(abs(mat)))

# Compute maximum row sum
max_row_sum <- max(rowSums(abs(mat)))

# Print the results
cat("Frobenius Norm:", frobenius_norm, "\n")
cat("Maximum Column Sum:", max_col_sum, "\n")
cat("Maximum Row Sum:", max_row_sum, "\n")
```


6. Solve the system of equations given in no. 4, starting with the initial vector of $[0,0,0]$ 
Since the system in no.4 is not diagonally dominant, thus not working with Jacobi or Seidel, we will use this system instead:
$$
\begin{gathered}
26x + 2y + 2z = 12.6 \\
3x + 27y + z = – 14.3 \\
2x + 3y + 17z = 6.0
\end{gathered}
$$
a. Solve using the Jacobi method with 2-digit precision.
```{r}
# Define the coefficients matrix (A)
A <- matrix(c(26, 2, 2, 3, 27, 1, 2, 3, 17), nrow = 3, byrow = TRUE)

# Define the constants vector (b)
b <- c(12.6, -14.3, 6)
z <- solve(A,b)
z <- round(z, digits = 2)
# True Solution
print(z)
# Define L, D, U
L=matrix(c(0,0,0,3,0,0,2,3,0), nrow = 3, byrow = T)
D=matrix(c(26,0,0,0,27,0,0,0,17), nrow = 3, byrow = T)
U=matrix(c(0,2,2,0,0,1,0,0,0), nrow = 3, byrow = T)
A==L+D+U
x0=matrix(c(0,0,0), nrow = 3, byrow = T)
# 1st Iteration
x1=(-1*inv(D))%*%(L+U)%*%x0+inv(D)%*%b
x1
# 2nd Iteration
x2=(-1*inv(D))%*%(L+U)%*%x1+inv(D)%*%b
x2
# 3rd Iteration
x3=(-1*inv(D))%*%(L+U)%*%x2+inv(D)%*%b
x3
# 4th Iteration
x4=(-1*inv(D))%*%(L+U)%*%x3+inv(D)%*%b
x4
# Last Iteration
x5=(-1*inv(D))%*%(L+U)%*%x4+inv(D)%*%b
x5
```

b. Solve using Gauss-Seidel method with 2-digit precision.
```{r}
# Define the coefficients matrix (A)
A <- matrix(c(26, 2, 2, 3, 27, 1, 2, 3, 17), nrow = 3, byrow = TRUE)

# Define the constants vector (b)
b <- c(12.6, -14.3, 6)
z <- solve(A,b)
z <- round(z, digits = 2)
# True Solution
print(z)
# Define L, D, U
L=matrix(c(0,0,0,3,0,0,2,3,0), nrow = 3, byrow = T)
D=matrix(c(26,0,0,0,27,0,0,0,17), nrow = 3, byrow = T)
U=matrix(c(0,2,2,0,0,1,0,0,0), nrow = 3, byrow = T)
A==L+D+U
x0=matrix(c(0,0,0), nrow = 3, byrow = T)
# 1st Iteration
x1=(-1*inv(L+D))%*%U%*%x0+inv(L+D)%*%b
x1
# 2nd Iteration
x2=(-1*inv(L+D))%*%U%*%x1+inv(L+D)%*%b
x2
# 3rd Iteration
x3=(-1*inv(L+D))%*%U%*%x2+inv(L+D)%*%b
x3
# Last Iteration
x4=(-1*inv(L+D))%*%U%*%x3+inv(L+D)%*%b
x4
```
c. Solve for $\bar{e}$ if the true solution is $x=(1.5,0.33,0.45)^T$.

```{r}
A <- matrix(c(26, 2, 2, 3, 27, 1, 2, 3, 17), nrow = 3, byrow = TRUE)
b <- c(12.6, -14.3, 6)
x_bar=matrix(c(1.5,0.33,0.45))
r=b-A%*%x_bar
e_bar=inv(A)%*%r
abs(e_bar)
# To check if x_bar + e_bar gives true solution
x_barcorrected=x_bar+e_bar
x_barcorrected
```
Thus, $\bar{e}$ is $(1,0.93,0.05)^T$

