---
title: "MAT 5101 - Activity Trapezoidal, Simpson, and Romberg Integration Rules"
subtitle: "Group 2"
author: "Erik Celdran, Jade Rosales, Marc Valeros"
date: "2024-04-02"
output: pdf_document
header-includes: |
  \usepackage{graphicx}
  \graphicspath{ {figures/} }
  \usepackage{array}
  \usepackage{amsmath}
  \usepackage{xcolor}
  \usepackage{bigints}
  \graphicspath{ {./} }
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(PolynomF)
library(matrixcalc)
library(pracma)
library(Deriv)
```

Estimate $sin(x^2)$ using Trapezoidal, Simpson and Romberg's integration rule. 


1. Using Trapezoidal Rule: 


2. Using Simpson's Rule:

Use Simpson's rule to estimate $\int_1^2 {sin(x^2)} d x$. Accurate within 0.001.
The formula for approximating areas using Simpson's rule is defined as follows:
$$
T_n=\frac{\Delta x}{3}\left(f\left(x_0\right)+4 f\left(x_1\right)+2 f\left(x_2\right)+4 f\left(x_3\right)+2 f\left(x_4\right)+\ldots+f\left(x_n\right)\right)
$$
where:
$$
\Delta x=\frac{b-a}{n} ; T_n \approx \int_a^b f(x) d x
$$

To identify the minimum number of iterations $n$, we use the error formula for the Simpson's rule:
$$
\bar{e} \leq \frac{M(b-a)^5}{180 n^4}
$$

Where $\mathrm{M}$ is the maximum value of $\left|f^{(4)}(x)\right|$ in the interval $(a, b)$. Rearranging the variables gives
$$
n \geq \sqrt[4]{\frac{M(b-a)^5}{180 \bar{e}}}
$$

To find $M$, we first get $f^{(4)}(x)$.

```{r} 

func = function(x) {sin(x^2)}
dp4 = Deriv(Deriv(Deriv(Deriv(func,"x"),"x"),"x"),"x") # 4th derivative

dp4(c(1,2))

```
So we set $M = 59.16023$ and proceed to solve for $n$.
$n \geq \sqrt[4]{\frac{59.16023(2-1)^5}{180(0.001)}}=4.25784 \Rightarrow 5$ iterations

```{r echo=FALSE}
n = (59.16023*(2-1)^5/(180*0.001))^(1/4)
n

func = function(x){ sin(x^2) }
df = data.frame(matrix(ncol=6))
colnames(df) = c("n", "x_n", "f(x_n)", "f or 2f or 4f(x_n)", "sum", "sum*delta/3")
count = 0
simpson = function(f, a, b, n) {
delta = (b-a)/n
x = a
sum = f(x)
df[nrow(df)+1, ] <<- list(count, x, f(x), f(x), sum, sum*delta/3) # f(x_0)
count <<- count + 1
x = x + delta
while(x < b) {
if(count %% 2 == 0) { temp = 2*f(x) }
else { temp = 4*f(x) }
sum = sum + temp
df[nrow(df)+1, ] <<- list(count, x, f(x), temp, sum, sum*delta/3) # f(x_1) to f(x_n-1)
count <<- count + 1
x = x + delta
}
sum = sum + f(x)
df[nrow(df)+1, ] <<- list(count, x, f(x), f(x), sum, sum*delta/3) #f(x_n)
df <<- df[-1,]
return (sum / 3 * delta)
}
answer = simpson(func, 1, 2, 6)
print(df, row.names = FALSE)

answer

actual =integrate(func,1,2)
abs(actual$value-answer) # absolute error 
```
Therefore, the answer is 0.4948502. The estimation of $sin(x^2)$ with 5 iterations of Simpson Integration yielded an absolute error of 0.0003420327




3. Using Romberg's Rule:

```{r} 

# Define the function

f <- function(x) {
  sin(x^2)
}


# Romberg integration function
romberg_function <- function(f, a, b, n = 5) {
  deltax <- (b - a) / 2
  T <- (deltax / 2) * (f(a) + 2*f(a+deltax) + f(b))
  result <- matrix(NA, nrow = n, ncol = n)
  result[1, 1] <- T
  
  for (i in 2:n) {
    deltax <- deltax / 2
    sum <- 0
    for (k in 1:(2^(i - 1))) {
      sum <- sum + f(a + (2 * k - 1) * deltax)
    }
    T <- 0.5 * T + deltax * sum
    result[i, 1] <- T
    for (j in 2:i) {
      result[i, j] <- (4^(j-1) * result[i, j - 1] - result[i - 1, j - 1]) / (4^(j - 1) - 1)
    }
  }
  return(result)
}

# Call the romberg function 
result <- romberg_function(f, 1,2)  
int = as.numeric(result[nrow(result), ncol(result)])



# Print the result
print(result)

#Check the pracma romberg function
romberg(f,1,2)

actual = integrate(f,1,2)

abs(actual$value-int)

```
The estimation of $sin(x^2)$ With 5 iterations of Romberg Integration yielded an absolute error of 1.828787e-11 or 0.00000000001828787
